Bound within Boundaries: How Well Do Protected Areas Match Movement Corridors of Their Most Mobile Protected Species?

### 2020-10-03 Chapter_0/04_Manuscript/Manuscript.tex
Finalized revision

\noindent 2. Between 2011 and 2019, we collected high-resolution GPS movement
data on 16 dispersing African wild dog (\textit{Lycaon pictus}) coalitions from
a free-ranging population in the Kavango-Zambezi Transfrontier Conservation Area
(KAZA-TFCA). Spanning five countries and 520'000 km\textsuperscript{2} the
KAZA-TFCA is the world's largest transboundary conservation area and a prime
example for international conservation efforts. We used integrated step
selection analysis to estimate habitat selection of dispersers and to create a
permeability surface for the entire KAZA-TFCA. We compared landscape
permeability across different regions within the KAZA-TFCA as well as outside
its boundaries. Lastly, we calculated least-cost paths and corridors to verify
that major movement routes were adequately encompassed within the KAZA-TFCA.

\noindent 3. Permeability within the boundaries of the KAZA-TFCA was more than
double compared to areas outside it. Furthermore, we observed a five-fold
permeability difference among the five KAZA-TFCA countries. We also showed
that major movement corridors of wild dogs ran within the KAZA-TFCA, although
some minor routes remained outside formally protected areas.

\noindent 4. Differences in permeability were mainly related to different
degrees of human activities across regions, and to the presence or absence of
rivers, swamps or open water. The relationship between permeability and other
landscape features was less pronounced.

@@

In recent years, a growing body of research has used animal relocation data to
identify movement corridors and assess connectivity at large scales (e.g.
\citealp{Chetkiewicz.2006, Squires.2013, Elliot.2014}). Identification of
potential movement corridors typically relies on the estimation of permeability
surfaces, which return the ease or willingness at which the focal species
traverses a specific landscape \citep{Sawyer.2011}. Such surfaces are created
based on species' relative selection strengths \citep{Avgar.2017}, which can be
quantified using a suite of selection functions \citep{Zeller.2012}.
Specifically, selection strengths are estimated by comparing spatial covariates
(e.g. environmental and anthropogenic) at locations visited by the animal to the
same spatial covariates at locations avilable to the animal \citep{Zeller.2012}.
Importantly, selection functions require adequate landscape and relocation data
that are representative of the process being studied \citep{Diniz.2020}. Altough
selection during residence and dispersal may coincide \citep{Fattebert.2015}, it
appears that relocation data collected on dispersing individuals outperforms
data collected on resident individuals in the detection of large-scale movement
corridors \citep{Elliot.2014, Abrahms.2017, Diniz.2020}. Nevertheless, dispersal
data is inherently difficult to collect and remains scarce in the connectivity
literature \citep{Vasudev.2015}. As such, most permeability surfaces upon which
movement corridors are identified are created using relocation data collected on
resident individuals. This has likely limited our ability to meaningfully assess
the effectiveness of protected areas in securing functional connectivity for
their protected species.

@@

Across the KAZA-TFCA, the African wild dog (\textit{Lycaon pictus}) represents a
highly mobile and endangered flagship species for conservation efforts. Once
widespread across the entire Sub-Saharan continent, wild dogs have been widely
extirpated through human persecution, habitat destruction, and disease outbreaks
\citep{Woodroffe.2012}. As a result, the species has become one of Africa's most
endangered large carnivores, and currently only survives in small, spatially
scattered subpopulations \citep{Woodroffe.2012}. Within these subpopulations,
wild dogs form cooperative breeding packs of up to thirty individuals
\citep{Creel.2002}, whose social structure is strongly governed by the process
of dispersal \citep{McNutt.1996, Behr.2020}. Both males and females disperse
from their natal pack, either alone or in same-sex dispersing coalitions, and
search for unrelated mates and a suitable territory to settle
\citep{McNutt.1996, Cozzi.2020, Behr.2020}. During dispersal, wild dogs can
cover several hundred kilometers and cross international borders
\citep{Masenga.2016, Woodroffe.2019, Cozzi.2020}. Despite the importance of
dispersal for the long-term viability of this species, little empirical
information is available on habitat selection and potential movement barriers
during dispersal. The few studies that have collected dispersal data have shown
that dispersers quickly move over large distances, avoid human-dominated
landscapes and areas densely covered by trees, but prefer proximity to water
\citep{Masenga.2016, Woodroffe.2019, Oneill.2020, Cozzi.2020}.

@@

\subsection{Permeability Surface}
Using the most parsimonious habitat selection model, we predicted a permeability
surface spanning the entire extent of the KAZA-TFCA. That is, we applied
\ref{EQ1} to our spatial covariates and calculated the selection score \(w(x)\)
for each raster cell. Because our representation of water was dynamic we
collapsed all dynamic water maps into a single map using areas that were covered
by water in at least 10\% of the cases. We used the resulting map to also
calculate a layer returning the distance to water. To reduce the influence of
outliers in predicted permeability scores, we followed \cite{Squires.2013} and
curtailed predicted scores between the 1\textsuperscript{st} and
99\textsuperscript{th} percentile of their original values. To compare
permeability across different regions, we normalized the permeability surface to
a range between 0 (most impermeable) and 1 (most permeable), and we determined
median permeability within and outside the KAZA-TFCA, within and outside
formally protected areas, and within each of the five KAZA-TFCA countries.

@@

\subsection{Habitat Selection Model}
Our most parsimonious habitat selection model (\(\Delta AIC > 2\) than any
alternative model; Table S4) retained the covariates \textit{water},
\textit{distance to water}, \textit{trees}, \textit{shrubs/grassland}, and
\textit{human influence}, beside the fixed covariates \textit{cos(ta)},
\textit{sl}, and \textit{log(sl)} (\Cref{PermeabilityResults}a). With respect to
environmental covariates, dispersers avoided moving through water (\(\beta\) =
-0.53, 95\% CI -0.79 to -0.27) but selected for locations in its vicinity,
although the latter effect was not significant (\(\beta\) = -0.33, 95\% CI =
-0.73 to 0.08). Dispersers avoided areas that were densely covered by trees
(\(\beta\) = -0.31, CI = -0.47 to -0.15) and preferred areas covered by
shrubs/grassland (\(\beta\) = 0.25, 95\% CI = 0.07 to 0.42). Finally, dispersers
avoided areas that were influenced by humans (\(\beta\) = -0.45, 95\% CI = -0.82
to -0.08). With the exepction of \textit{distance to water}, random effects
revealed little variability between dispersal coalitions (Appendix A6).

% The positive coefficient for \(cos(ta)\) (\(\beta\) = 0.10; 95\% CI = 0.05 to
% 0.15) suggested that the uniform distribution for turning angles produced too
% little directionality. Similarly, the positive coefficient for \(log(sl)\)
% (\(\beta\) = 0.12, 95\% CI = 0.02 to 0.22) indicated that our fitted gamma
% distribution produced step-lengths slightly shorter than the true
% selection-free gamma distribution would have. The estimate for \(sl\), on the
% other hand, was insignificant (\(\beta\) = 0.03, 95\% CI = -0.09 to 0.15).

@@

Our approach of identifying movement corridors based on pre-defined start and
end points implicitly assumes that individuals know the end point of their
dispersal journey and that they have almost complete knowledge of associated
movement costs \citep{Panzacchi.2016}. Since dispersers often move into unknown
territory, this may not necessarily be the case \citep{Abrahms.2017,
Cozzi.2020}. However, specification of pre-defined end points might not be
necessary, as the parametrized iSSF model can be used as mechanistic movement
model to simulate dispersal events from known source points, yet without
restricting the domain of potential end points \citep{Signer.2017}.
Consequently, movement corridors would emerge more naturally as the result of a
myriad of simulated dispersal events \citep{Zeller.2020}.

Besides estimating corridors, individual based simulations may be used to
generate permeability surfaces \citep{Avgar.2016, Signer.2017}. Such surfaces
have been shown to reduce the risk of overestimating \(w(x)\) and conesquently
connectivity, particularly in areas that lie far from suitable habitats
\citep{Signer.2017}. While a simulation-based approach is conceptually
straightforward, computational requirements for such a large spatial extent as
the KAZA-TFCA are very high, making the use of this approach challenging. We
urge future studies to optimize the simulation of corridors from iSSFs that
capture a more mechanistic model of dispersal.

>>>>

\noindent 2. Between 2011 and 2019, we collected high-resolution GPS data on 16
dispersing African wild dog (\textit{Lycaon pictus}) coalitions from a
free-ranging population in the Kavango-Zambezi Transfrontier Conservation Area
(KAZA-TFCA). Spanning five countries and 520'000 km\textsuperscript{2} the
KAZA-TFCA is the world's largest transboundary conservation area and a prime
example for international conservation efforts. We used integrated step
selection analysis to estimate habitat selection of dispersers and to create a
permeability surface for the KAZA-TFCA. We compared landscape permeability
across different regions within the KAZA-TFCA as well as outside its boundaries.
Lastly, we calculated least-cost paths and corridors to verify that major
movement routes were adequately encompassed within the KAZA-TFCA.

\noindent 3. Permeability within the boundaries of the KAZA-TFCA was more than
double compared to areas outside it. Furthermore, we observed a five-fold
permeability difference among the five KAZA-TFCA countries. We also showed that
major movement corridors of wild dogs ran within the KAZA-TFCA, although some
minor routes remained formally unprotected.

\noindent 4. Differences in permeability were mainly related to different
degrees of human activities across regions, and to the presence or absence of
rivers, swamps and open water. The relationship between permeability and other
landscape features was less pronounced.

@@

In recent years, a growing body of research has used animal relocation data to
identify movement corridors and assess connectivity at large scales (e.g.
\citealp{Chetkiewicz.2006, Squires.2013, Elliot.2014}). Identification of
potential movement corridors typically relies on the estimation of permeability
surfaces, which return the ease or willingness at which the focal species
traverses a specific landscape \citep{Sawyer.2011}. Such surfaces are created
based on species' relative selection strengths \citep{Avgar.2017}, which can be
quantified using a suite of selection functions \citep{Zeller.2012}.
Specifically, selection strengths are estimated by comparing spatial covariates
(e.g. environmental and anthropogenic) at locations visited by the animal to the
same spatial covariates at locations available to the animal
\citep{Zeller.2012}. Importantly, selection functions require adequate landscape
and relocation data that are representative of the process being studied
\citep{Diniz.2020}. Altough selection during residence and dispersal may
coincide \citep{Fattebert.2015}, it appears that relocation data collected on
dispersing individuals outperforms data collected on resident individuals in the
detection of large-scale movement corridors \citep{Elliot.2014, Abrahms.2017,
Diniz.2020}. Nevertheless, dispersal data is inherently difficult to collect and
remains scarce in the connectivity literature \citep{Vasudev.2015}. As such,
most permeability surfaces upon which movement corridors are identified are
created using relocation data collected on resident individuals. This has likely
limited our ability to meaningfully assess the effectiveness of protected areas
in securing connectivity for their protected species.

@@

Across the KAZA-TFCA, the African wild dog (\textit{Lycaon pictus}) represents a
highly mobile and endangered flagship species for conservation efforts. Once
widespread across the entire Sub-Saharan continent, wild dogs have been widely
extirpated through human persecution, habitat destruction, and disease outbreaks
\citep{Woodroffe.2012}. For these reasons, viable populations mainly occur in
spatially scattered subpopulations within protected areas \citep{Woodroffe.1999,
Woodroffe.2012, VanDerMeer.2014}. Within these subpopulations, wild dogs form
cooperative breeding packs of up to thirty individuals \citep{Creel.2002}, whose
social structure is strongly governed by the process of dispersal
\citep{McNutt.1996, Behr.2020}. Both males and females disperse from their natal
pack, either alone or in same-sex dispersing coalitions, and search for
unrelated mates and a suitable territory to settle \citep{McNutt.1996,
Cozzi.2020, Behr.2020}. During dispersal, wild dogs can cover several hundred
kilometers and cross international borders \citep{Masenga.2016, Woodroffe.2019,
Cozzi.2020}. Despite the importance of dispersal for the long-term viability of
this species, little empirical information is available on habitat selection and
potential movement barriers during dispersal. The few studies that have
collected dispersal data have shown that dispersers quickly move over large
distances, avoid human-dominated landscapes and areas densely covered by trees,
but prefer proximity to water \citep{Masenga.2016, Woodroffe.2019, Oneill.2020,
Cozzi.2020}.

@@

\subsection{Permeability Surface}
Using the most parsimonious habitat selection model, we predicted a permeability
surface spanning the entire extent of the KAZA-TFCA. That is, we applied
\ref{EQ1} to our spatial covariates and calculated the selection score \(w(x)\)
for each raster cell. Because our representation of water was dynamic we
collapsed all dynamic water maps into a single map using areas that were covered
by water in at least 10\% of the cases. We used the resulting map to also
calculate a layer returning the distance to water. Because the delta only covers
5\% of the KAZA-TFCA, we considered the use of a single water map to be
appropriate. To reduce the influence of outliers in predicted permeability
scores, we followed \cite{Squires.2013} and curtailed predicted scores between
the 1\textsuperscript{st} and 99\textsuperscript{th} percentile of their
original values. To compare permeability across different regions, we normalized
the permeability surface to a range between 0 (most impermeable) and 1 (most
permeable), and we determined median permeability within and outside the
KAZA-TFCA, within and outside formally protected areas, and within each of the
five KAZA-TFCA countries.

@@

\subsection{Habitat Selection Model}
Our most parsimonious habitat selection model (\(\Delta AIC > 2\) than any
alternative model; Table S4) retained the covariates \textit{water},
\textit{distance to water}, \textit{trees}, \textit{shrubs/grassland}, and
\textit{human influence}, beside the fixed covariates \textit{cos(ta)},
\textit{sl}, and \textit{log(sl)} (\Cref{PermeabilityResults}a). Dispersers
avoided moving through water (\(\beta\) = -0.53, 95\% \(CI\) -0.79 to -0.27) but
selected for locations in its vicinity, although the latter effect was not
significant (\(\beta\) = -0.33, 95\% \(CI\) = -0.73 to 0.08). Dispersers avoided
areas that were densely covered by trees (\(\beta\) = -0.31, \(CI\) = -0.47 to
-0.15) and preferred areas covered by shrubs/grassland (\(\beta\) = 0.25, 95\%
\(CI\) = 0.07 to 0.42). Finally, dispersers avoided areas that were influenced
by humans (\(\beta\) = -0.45, 95\% \(CI\) = -0.82 to -0.08). Except for
\textit{distance to water} (\(SD\)\textsubscript{RandomEffect} = 0.57), we
observed little variation between dispersal coalitions' relative selection
strengths (\(SD\)\textsubscript{RandomEffect} \(<\) 0.22 for all other
covariates, see also Figure S7).

@@

Our approach of identifying movement corridors based on pre-defined start and
end points implicitly assumes that individuals know the end point of their
dispersal journey and that they have almost complete knowledge of associated
movement costs \citep{Panzacchi.2016}. Since dispersers often move into unknown
territory, this may not necessarily be the case \citep{Abrahms.2017,
Cozzi.2020}. However, specification of pre-defined end points might not be
necessary, as the parametrized iSSF model can be used as mechanistic movement
model to simulate dispersal from known source points, yet without restricting
the domain of potential end points \citep{Signer.2017}. Consequently, movement
corridors would emerge more naturally as the result of a myriad of simulated
dispersal events \citep{Allen.2016, Zeller.2020}.

Besides estimating corridors, individual based simulations may be used to
generate permeability surfaces \citep{Avgar.2016, Signer.2017}. Such simulation
based surfaces have been shown to reduce the risk of overestimating permeability
(\(w(x)\)) and conesquently connectivity, particularly in areas that lie far
from suitable habitats \citep{Signer.2017}. While a simulation-based approach is
conceptually straightforward, computational requirements for such a large
spatial extent as the KAZA-TFCA are very high, making the use of this approach
challenging. We therefore urge future studies to optimize the simulation of
movement from iSSFs to capture a more mechanistic model of dispersal.
